---
title: "Mapping the Metro Manila Subway"
description: ""
date: 2022-11-08
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA, message=FALSE, warning=FALSE)
```

Today I'll experiment with making maps via `leaflet`, which I'm using for the first time. I'm relying mainly on [this tutorial](https://rstudio.github.io/leaflet).

```{r}
library(tidyverse)
library(leaflet)
```

Below is a map pointing out some of the planned stations of the future Metro Manila Subway, from Quirino Highway to 11th Avenue in BGC. I got the locations from trawling through news articles and project documents, then I used Google Maps to get their coordinates.

```{r}
leaflet(options = leafletOptions(minZoom = 10, maxZoom = 15)) %>%
  addTiles() %>%
  addMarkers(lng=121.028460, lat=14.689541, label="Quirino Highway Station") %>%
  addMarkers(lng=121.032355, lat=14.676936, label="Tandang Sora Station") %>%
  addMarkers(lng=121.035685, lat=14.654850, label="North Avenue Station") %>%
  addMarkers(lng=121.037591, lat=14.644747, label="Quezon Avenue Station") %>%
  addMarkers(lng=121.051628, lat=14.640692, label="East Avenue Station") %>%
  addMarkers(lng=121.065282, lat=14.627151, label="Anonas Station") %>%
  addMarkers(lng=121.069868, lat=14.613690, label="Camp Aguinaldo Station") %>%
  addMarkers(lng=121.063565, lat=14.588103, label="Ortigas Station") %>%
  addMarkers(lng=121.061238, lat=14.575162, label="Shaw Station") %>%
  addMarkers(lng=121.055859, lat=14.558327, label="11th Avenue Station")
```

And here's the map! It's fine? It's a little busy, so let's change the map tile from the default OpenStreetMap to a nice minimalist one from [this list](http://leaflet-extras.github.io/leaflet-providers/preview/index.html). 

```{r, preview=TRUE}
leaflet(options = leafletOptions(minZoom = 10, maxZoom = 15)) %>%
  addProviderTiles(providers$CartoDB.Voyager) %>%
  addMarkers(lng=121.028460, lat=14.689541, label="Quirino Highway Station") %>%
  addMarkers(lng=121.032355, lat=14.676936, label="Tandang Sora Station") %>%
  addMarkers(lng=121.035685, lat=14.654850, label="North Avenue Station") %>%
  addMarkers(lng=121.037591, lat=14.644747, label="Quezon Avenue Station") %>%
  addMarkers(lng=121.051628, lat=14.640692, label="East Avenue Station") %>%
  addMarkers(lng=121.065282, lat=14.627151, label="Anonas Station") %>%
  addMarkers(lng=121.069868, lat=14.613690, label="Camp Aguinaldo Station") %>%
  addMarkers(lng=121.063565, lat=14.588103, label="Ortigas Station") %>%
  addMarkers(lng=121.061238, lat=14.575162, label="Shaw Station") %>%
  addMarkers(lng=121.055859, lat=14.558327, label="11th Avenue Station")
```

Neat. I'll revisit this later on to see what else I can add.